NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2020-11-30 Manuscript/manuscriptPGPE.tex
Update manuscript

To characterize the differences in lipid headgroup conformational ensembles
leading to the distinct order parameters in PS and PG lipids, we calculate the
the distributions of heavy atom dihedral angles from CHARMM36 simulations.
Among available simulation models, this force field has least problems in
reproducing the experimental lipid headgroup and glycerol backbone order parameters in different lipids,
see Refs. \cite{botan15} and \cite{antila19}, and Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}.
Importantly, it captures the experimentally observed distinct order parameters for PS and PG headgroups (Fig. \ref{??}).

show major differences between lipids only for O$_\alpha$-C$_\alpha$ and  C$_\alpha$-C$_\beta$ bonds
near the end of the headgroup (Fig. \ref{structures}).
These probably explain the slight differences in the
angle between headgroup dipole and membrane normal, which decreases 
in the order PG $>$ PE  $>$ PC  $>$ PS, although difference between PC and PE may an artefact as
the $\beta$-carbon order parameter in PC is poorly reproduced by the CHARMM36 force field.
As in previous NMRlipids project results for PC and PS lipids \cite{botan15,antila19},
none of the MD simulation force fields correctly captures all the
headgroup and glycerol backbone order parameters of PE and PG lipids
(Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG})
that would enable a straightforward interpretation of conformational ensembles.
%show wide variation between different force fields
%and none of the force fields reproduce all values within experimental error bars.
%(Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}),
%as observed previously also for PC and PS lipids \cite{botan15,antila19}.
%The Slipid simulations were able to capture the essential differences between PC and PS lipid
%headgroups \cite{antila19}, but this is not the case for PE and PG lipids.
%Without further discussion about poorly performing force fields, we focus on more detailed analysis
%of
Nevertheless, CHARMM36 force field, which gives the results closest to the experiments for all lipids,
captures the essential differences between
PC, PS, PG and PE headroup order parameters  (Fig.~\ref{structures}) with the exception of $\beta$-carbon order parameter of PC
%give the best overall agreement with experiments for
%the headgroup and glycerol backbone order parameters of all PC, PS, PG and PE lipids in this and our previous
%studies \cite{botan15,antila19}. Even thought many order parameters in CHARMM36 simulations are
%not within the experimental error bars, it 
which is too negative when compared with other lipids or experiments \cite{botan15}.
% As already pointed out in our previous study \cite{antila19},

>>>>

To characterize the differences in lipid headgroup conformational ensembles
leading to the distinct order parameters in PS and PG lipids, we calculate the
the distributions of heavy atom dihedral angles from CHARMM36 simulations.
Among available simulation models, this force field has least problems in
reproducing the experimental lipid headgroup and glycerol backbone order parameters in different
lipids \cite{botan15,antila19} (Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}).
Importantly, it captures the experimentally observed distinct order parameters for PS and PG headgroups (Fig. \ref{??}).
The dihedral distributions show significant similarity between different lipids for all the other
bonds, except for the last two in the headgroup end, O$_\alpha$-C$_\alpha$ and  C$_\alpha$-C$_\beta$ bonds,
in figure \ref{structures}. Therefore, we conclude that the disctinct order parameters of
PS and PG headgroups can be explained by the differences in these two dihedrals with respenct to other lipids.
The differences between PC and PE lipids, especially in O$_\alpha$-C$_\alpha$ bond dihedral,
may be an artefact because the $\beta$-carbon order parameter in PC is poorly reproduced by the CHARMM36 force field \cite{botan15}.
In larger scale, these differences are reflected, for example, to the 
angle between headgroup dipole and membrane normal,
which decreases in the order of PG $>$ PE  $>$ PC  $>$ PS (Fig. \ref{structures}).

%
%As in previous NMRlipids project results for PC and PS lipids \cite{botan15,antila19},
%none of the MD simulation force fields correctly captures all the
%headgroup and glycerol backbone order parameters of PE and PG lipids
%(Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG})
%that would enable a straightforward interpretation of conformational ensembles.
%
%show wide variation between different force fields
%and none of the force fields reproduce all values within experimental error bars.
%(Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}),
%as observed previously also for PC and PS lipids \cite{botan15,antila19}.
%The Slipid simulations were able to capture the essential differences between PC and PS lipid
%headgroups \cite{antila19}, but this is not the case for PE and PG lipids.
%Without further discussion about poorly performing force fields, we focus on more detailed analysis
%of
%
%Nevertheless, CHARMM36 force field, which gives the results closest to the experiments for all lipids,
%captures the essential differences between
%PC, PS, PG and PE headroup order parameters  (Fig.~\ref{structures}) with the exception of $\beta$-carbon order parameter of PC
%
%give the best overall agreement with experiments for
%the headgroup and glycerol backbone order parameters of all PC, PS, PG and PE lipids in this and our previous
%studies \cite{botan15,antila19}. Even thought many order parameters in CHARMM36 simulations are
%not within the experimental error bars, it 
%
%which is too negative when compared with other lipids or experiments \cite{botan15}.
%
% As already pointed out in our previous study \cite{antila19},
